---
title: "Integrative annotation for human genetics: Approaches with Bioconductor"
author: "Vincent J. Carey"
date: "`r doc_date()`"
abstract: >
  The study of genetic variation among human individuals
  engenders substantial archives of details on individual genomes.
  Details of genetic variation are interpreted using evolving
  resources assembled at various scales.  NCBI's dbSNP assembles
  information on relatively small variants that occur with
  some regularity in human populations.  The EMBL-EBI GWAS catalog
  provides a curated view of replicated studies of genotype-phenotype
  association.  This workflow reviews the use of Bioconductor packages
  that help investigators interpret their own findings in studies
  of human genetics.
vignette: >
  %\VignetteIndexEntry{Integrative annotation for human genetics: Approaches with Bioconductor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document
---

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(gwascat)
library(GenomicFiles)
library(rtracklayer)
library(AnnotationHub)
library(DT)
library(SIFT.Hsapiens.dbSNP132)
library(grasp2db)
library(BSgenome)
library("SNPlocs.Hsapiens.dbSNP144.GRCh37")
library(BSgenome.Hsapiens.NCBI.GRCh38)
})
```

# Background

The table of contents of Vogel and Motulsky's [_Human Genetics: Problems and Approaches_](https://books.google.com/books?id=xuztCAAAQBAJ&lpg=PA6&dq=human%20genetics&pg=PR32#v=onepage&q=human%20genetics&f=false)
is a worthy survey of concepts addressed in research on human
genetics and genetic medicine.  The frontiers of knowledge in the field
are shifting, and expectations are high.

In this workflow, I aim to show how researchers can use R
to interrogate important resources of use in human genetic
epidemiology and medical genomics.  I show how to program with two
genome-wide association study
(GWAS) catalogs, 
the [EMBL-EBI GWAS catalog](https://www.ebi.ac.uk/gwas/) 
and the [NHLBI GRASP v2.0](http://iapps.nhlbi.nih.gov/GRASP/Overview.aspx).
Aspects of findings reported in these studies are then
integrated with new functional and structural annotation resources
to aid in variant interpretation.
An appendix provides
brief treatment of "reference genome builds" for _Homo sapiens_,
packages for querying contents of
the [NCBI dbSNP](http://www.ncbi.nlm.nih.gov/SNP/),
and tools for obtaining and programming with gene models.

# The gwascat package for the EMBL-EBI GWAS catalog

## Basic operations, fields, and interactive tabulation 

The EMBL-EBI catalog is presented using GRCh38 coordinates.
```{r lkg}
library(gwascat)
data(ebicat38)
ebicat38
```

While there are `r length(ebicat38)` records, the number of unique loci is
```{r lku}
length(unique(ebicat38$SNPS))
```

A full view of the metadata about each study result is available
with the commands
```{r eval=FALSE}
library(DT)
datatable(as.data.frame(mcols(ebicat38)), options=list(autoWidth=TRUE,
  style="height:30px"), pageLength=5)
```

The following command generates
a table restricting attention to records related to asthma.

```{r gett}
suppressWarnings({
aind = grep("sthma", ebicat38$DISEASE.TRAIT)
datatable(as.data.frame(mcols(ebicat38[aind])), options=list(autoWidth=TRUE,
  style="height:30px", pageLength=5))
})
```

## Navigating traits using the EMBL-EBI Experimental Factor Ontology

Field `MAPPED_TRAIT_URI` includes a comma-delimited string with
URIs referring to an ontology for traits and other factors relevant
to biological experiments and observations.  The underlying
ontology is available in the form of an annotated algebraic graph.

```{r lkggg}
data(efo.obo.g)
efo.obo.g
```
There are over 16000 terms in the ontology.  Terms and term-related
metadata are manipulated using methods of the `r Biocpkg("graph")`
package.
```{r lkgt}
nodes(efo.obo.g)[1:4] # imported directly from OBO
names(nodeData(efo.obo.g)[[1]])
sapply(nodeData(efo.obo.g)[1:4], "[[", "name")
```

Let's obtain the EFO annotation for SNP `rs347412`.
```{r lkefo}
ind = which(ebicat38$SNPS == "rs347412")
urs = ebicat38$MAPPED_TRAIT_URI[ind]
urs
```
These entries must be converted to match the EFO OBO node
nomenclature.  We then find the EFO names of the factors annotated
to this SNP.
```{r lkefo2}
nn = uri2node(urs)
nd = nodeData(efo.obo.g, nn)
sapply(nd, "[[", "name")
```

The current representation of the ontology is a directed graph
with links pointing from a term to its semantic parent.  We
convert to an undirected graph to explore semantic neighborhoods of terms.
The `adj` method will return the nodes adjacent to a specified node.
Here we obtain the terms accessible from `respiratory system disease`
with a single step.

```{r explsem}
rsdn = adj(ugraph(efo.obo.g), "EFO:0000684")  # respiratory system disease
unlist(sapply(nodeData(efo.obo.g, rsdn[[1]]), "[[", "name"))
```

The `r Biocpkg("RBGL")` package can be used to deploy diverse graph algorithms
against this ontology.

Once a node name of interest has been found, `node2uri` can be used
with code to find
GWAS hits deemed relevant by the curators.  We'll work with hg19
coordinates.

```{r lkuri}
data(ebicat37)
library(GenomeInfoDb)
seqlevelsStyle(ebicat37) = "UCSC"
genome(ebicat37) = "hg19"
e270 = ebicat37[ grep(node2uri("EFO:0000270"), ebicat37$MAPPED_TRAIT_URI) ]
length(e270)
table(e270$DISEASE.TRAIT)[1:5]
```

# GRASP

GRASP is a much denser catalog requiring a different
approach to archiving and query resolution.  First
execution
of `GRASP2()` will trigger a download of a 5GB
SQLite database that can then be used with `r CRANpkg("dplyr")`
programming.  Therefore this document does not evaluate
the following chunk, but the output is precomputed and
left static.

```{r dogra,eval=FALSE}
library(grasp2db)
v = tbl(GRASP2(), 'variant')
v %>% filter(Phenotype == "Asthma")
```

<pre><code>## Source: sqlite 3.8.6 [AnnotationHub()[[&quot;AH21414&quot;]]]
## From: variant [33,351 x 33]
## Filter: Phenotype == &quot;Asthma&quot; 
## 
##        NHLBIkey     PMID HUPfield SNPid_dbSNP134 chr_hg19  pos_hg19
## 1    2086050316 20860503 1/1/2014             18        7  11597475
## 2   20860503866 20860503 1/1/2014            535        9 138396251
## 3  208605031097 20860503 1/1/2014            686        5 174868700
## 4  208605031186 20860503 1/1/2014            699        1 230845794
## 5  208605031603 20860503 1/1/2014           1117        3  22085809
## 6  208605031980 20860503 1/1/2014           1320       22  22599537
## 7  208605032429 20860503 1/1/2014           1535       11  61597972
## 8  208605032734 20860503 1/1/2014           1695       11  67352689
## 9  208605032835 20860503 1/1/2014           1760        8    442079
## 10 208605033085 20860503 1/1/2014           1899       15  41689232
## ..          ...      ...      ...            ...      ...       ...
## Variables not shown: SNPidInPaper (chr), LocationWithinPaper (chr), Pvalue
##   (dbl), NegativeLog10PBin (int), Phenotype (chr), PlatformSNPsPassingQC
##   (chr), GWASancestryDescription (chr), InGene (chr), InLincRNA (chr),
##   InMiRNA (chr), InMiRNABS (chr), dbSNPfxn (chr), dbSNPMAF (chr),
##   dbSNPallelesHetSe (chr), dbSNPvalidation (int), dbSNPClinStatus (chr),
##   ORegAnno (chr), ConservPredTFBS (chr), HumanEnhancer (chr), RNAedit
##   (chr), PolyPhen2 (chr), SIFT (chr), LS_SNP (chr), UniProt (chr),
##   EqtlMethMetabStudy (int), DiscoverySampleDescription (chr),
##   ReplicationSampleDescription (chr)</code></pre>

# Genomic contexts and interpretations of variants

## Presence in exons

We can map our GWAS hits to exons using the TxDb infrastructure.
```{r lkexoh}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
allex = exons(TxDb.Hsapiens.UCSC.hg19.knownGene)
subsetByOverlaps( e270, allex )
```

## SIFT scores

We query the SIFT resource using dbSNP identifiers.

```{r lksift}
rsids = e270$SNPS
library(SIFT.Hsapiens.dbSNP132)
subst = c("RSID", "METHOD", "PREDICTION", "SCORE")
sif = AnnotationDbi::select(SIFT.Hsapiens.dbSNP132, keys=rsids, cols=subst)
datatable(na.omit(sif))
```

## ChromImpute segmentation

We'll use adult lung sample from the epigenomics road map
as stored in the `r Biocpkg("erma")` package.  Other segmentations
can be retrieved using the `r Biocpkg("AnnotationHub")` package.

```{r lkcimp}
library(erma)
ee = makeErmaSet()[,31] # lung
library(rtracklayer)
cstates = import(files(ee)[1], which=e270, genome="hg19")
sort(table(cstates$name), decreasing=TRUE)
```

## Regions of chromatin modification

We'll check for coincidence of our GWAS hits with peaks identified
with H3K4me1 marks in fetal lung fibroblasts, using component AH43875
of the `r Biocpkg("AnnotationHub")`.
```{r lkh3}
library(AnnotationHub)
ah = AnnotationHub()
h3kf = ah[["AH43875"]]
subsetByOverlaps(e270, h3kf)
```

# Conclusions

The use of `r Biocpkg("GenomicRanges")` infrastructure
for representing sets of DNA variants leads to fairly
simple merge and intersection operations based on genomic
coordinates.  These operations are useful for sorting variants
into categories based on structural or functional modeling.
Richly annotated ranges can be used to manage and program
with GWAS catalogs, leading to efficient coupling of genomic
assay results with findings of genetic epidemiology.

# Appendix: Bioconductor infrastructure supporting genetic data analysis 

## Reference builds of the human genome sequence

The most recent build of the human genomic sequence
is labeled GRCh38.  Using Bioconductor, we can be very concrete about what this
is.

```{r lkhg}
library(BSgenome.Hsapiens.NCBI.GRCh38)
class(Hsapiens)
Hsapiens
class(Hsapiens$"1")
Hsapiens$"1"
```

## From dbSNP to GRanges

A number of packages represent snapshots of NCBI dbSNP.
```{r lkbs}
library(BSgenome)
available.SNPs()
```
Functions available for a recent build are:
```{r lklk}
library("SNPlocs.Hsapiens.dbSNP144.GRCh37")
ls(pos="package:SNPlocs.Hsapiens.dbSNP144.GRCh37")
```
We can retrieve data on a chromosome.  Note the
peculiar nomenclature for chromosomes.
```{r lksn}
snpsBySeqname(SNPlocs.Hsapiens.dbSNP144.GRCh37, "ch20")
```
